% ============================================================================
% RELATED WORK — PRISM AAAI 2026 (Final Rewrite)
% Citation keys match aaai2026_english.bib exactly.
% ============================================================================

\section{Related Work}
\label{sec:related}

\textbf{Neural-symbolic constraint reasoning.} Early neural-symbolic work combined perception networks with symbolic programs~\citep{Mao2019,Yi2019}. \citet{Pan2023} propose Logic-LM, establishing a translate-execute-repair pipeline for LLM-based formal reasoning; \citet{Olausson2023} integrate Z3 for explicit constraint verification; \citet{Ye2023} explore formal-method integration in LLM reasoning pipelines. These systems improve over pure language model reasoning on structured tasks but exhibit two persistent deficiencies: coarse UNSAT feedback triggers stagnation cycles, and independent per-problem solving discards all accumulated experience. PRISM addresses both through formally-grounded paradigm libraries and typed repair memory.

\textbf{Agent memory and experience accumulation.} Memory-augmented agents accumulate task-level insights across trajectories~\citep{Shinn2023,Zhao2024,Ouyang2025}, demonstrating cross-task transfer in web navigation, embodied reasoning, and software engineering. ExpeL~\citep{Zhao2024} extracts unverified natural-language insights and maintains them with upvote/downvote operators. When applied to CSP reasoning, such frameworks face two limitations: (a) memory units are unverified prose, making incorrect heuristics indistinguishable from correct ones at storage time; (b) task-level granularity fails to capture the operation-level semantic patterns required for constraint reasoning (e.g., ``when constraint types $X$ and $Y$ coexist with domain property $P$, this inference step is sound''). PRISM overcomes both by using Z3 as a quality oracle for paradigms and recording repair events at individual step granularity.

\textbf{Formal methods and constraint solving.} Classical CSP solvers employ systematic search with constraint propagation~\citep{Dechter2003,Russell2021}. The Z3 SMT solver~\citep{DeMoura2008} provides decidability procedures for rich logical theories and serves as PRISM's verification oracle. PRISM's distinctive contribution is using Z3 not merely as a back-end solver but as a \textit{quality-screening oracle} for candidate experience—a role that enables empirical certification of solving strategies before storage, complementing rather than replacing Z3's solver function.

\textbf{Learning reusable inference rules.} PRISM is most closely related to the tradition of learning reusable inference patterns from experience. Explanation-Based Learning (EBL) generalizes successful traces into reusable schemas with provable correctness derived from a domain theory~\citep{MitchellKellerKedarCabelli1986,DeJongMooney1986}. Macro-operator learning compresses primitive operator sequences into reusable macros~\citep{KorfMacro1985}, with \citet{Minton1990} analyzing the utility tradeoff between coverage and matching cost. In the symbolic-solver tradition, conflict-driven clause learning~\citep{MarquesSilvaSakallah1996} and lemma generation in SMT/model checking~\citep{BradleyIC3} accumulate experience from failed branches and feed it back as new constraints; minimal unsatisfiable subset enumeration~\citep{LiberatoreSchaerf} underpins PRISM's UNSAT-core analysis.

PRISM occupies a specific position in this spectrum. Unlike EBL macros, our paradigms are induced from \textit{noisy} LLM-generated traces (not provably sound) and certified \textit{empirically} via sampling-based screening rather than first-principles proof from a hand-built domain theory. Unlike clause/lemma learning in classical SMT, PRISM operates at the level of \textit{reusable inference patterns across problem instances}, not conflict clauses scoped to a single solver invocation. To our knowledge, PRISM is the first to combine LLM-induced inference rules with SMT-based empirical screening, filling the gap between provably-sound EBL macros (limited by a hand-built domain theory) and unverified LLM-memory heuristics (limited by lack of mechanical quality control).
